Seminar: Programming models and Code Generation\\ Winter Term 2014/2015\\ {\bf Quasi deterministic parallel programming using LVars}

### 2014-12-26 draft.tex
Update
Latest

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically.  Written in a hypothetical parallel language. 
\begin{lstlisting}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
	_ = put lv 4  /write 4 in lv/
	  in v   /final value is v/
\end{lstlisting}
The \textit{let par} expression, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. \par 
\begin{lstlisting}
let par  _ = put l (4, )
         _ = put l (, 3) /error/
         in get l
\end{lstlisting}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

\textit{\textbf{Existing models}} A number of models currently exist as a part of the long running effort towards deterministic by construction parallel programming. \par

@@

In a directed graph where v represents a particular node, we wish to find all (or the first k degrees of) node connected to v, then analyze each node in that set. Existing parallel solutions [1] often use a non-deterministic traversal of the connected component even though the final connected component is deterministic. For example, IVars cannot accumulate sets of visited nodes, nor can they be used as “mark bits” on visited nodes, since they can only be written once and not tested for emptiness. Streams, on the other hand, impose an excessively strict ordering for computing the unordered set of vertex labels in a connected component. A purely functional attempt is given in the original paper[1] where connected-component discovery is parallel, but members of the output set do not become available to other computations until component discovery is finished, limiting parallelism. 
An LVars based programming model can possibly allow us to write a breadth first search traversal where threads communicate using shared monotonic data structures. Consumers of the data structure may execute as soon as data is available, but may only observe irrevocable, monotonic properties of it.

@@

To describe Lvar formally, a parallel call-by-value $\lambda$-calculus named $\lambdaLVar$ is introduced in the paper [1] which is extended with a $store$ and communication primitives $put$ and $get$ that operate on data in the $store$. The definition of $\lambdaLVar$ is parameterized by the choice of lattice and therefore $\lambdaLVar$ is actually a family of languages, rather than a single language. An LVar may have an arbitrary number of states forming a set $D$, which is partially ordered by a relation $\userleq$. Formally, $D$ is a \emph{bounded-join-semilattice} augmented with a greatest element $\top$.

% free floating figure using width of full page, to be put on [t]op.
%\begin{figure}
%\centerline{
%\includegraphics[height=2in,width=1.0\columnwidth]{Figures/try.jpg}
%}
%\caption{A nice caption. The larger width allows for more text without
%taking too much space.}
%\label{Fig2}
%\end{figure}
\begin{itemize}
\item $D$ has a least element $\bot$, representing the initial “empty” state of an Lvar,
\item $D$ has a greatest element $\top$, representing the “error” state that results from conflicting updates to an Lvar,
\item $D$ comes equipped with a partial order $\userleq$, where $\bot \userleq d \userleq \top$ for all $d \in D$.
\item Every pair of elements in $D$ has a least upper bound (lub) written $\sqcup$ which means it is possible for two sub-computations to independently update an LVar, and then deterministically merge the results by taking the lub of the resulting two states.
\end{itemize}
Data structures like arrays, tree, maps , streams etc. can be represented  as lattice. $Figure$ $2$ gives three examples of lattices for common data structures.

During the evaluation of a $\lambdaLVar$ program, a \emph{store} $S$ keeps track of the states of LVars. Each LVar is represented by a \emph{binding} from a location $l$, drawn from a countable set $\Loc$, to its \emph{state}, which is some element $d \in D$. Although each LVar in a program has its own state, the states of all the LVars are drawn from the same lattice $D$. We can do this with no loss of generality because lattices corresponding to different types of LVars could always be union ed into a single lattice (with shared $\bot$ and $\top$ Elements). \\ \\
\textbf{Definition 1}: A \emph{store} is either a finite partial mapping $S : \Loc \fmap (D - \setof{\top})$, or the distinguished element $\topS$. \\ \\
The state space of $stores$ forms a \emph{bounded-join-semilattice} augmented with a greatest element, just as $D$ does, with the empty store $\bot_S$ as its least element and $\top_S$ as its greatest element. The $\userleq$ and $\sqcup$ operations defined on elements of $D$ can be lifted to the level of stores.\\ \\
\textbf{Definition 2}: 
 A store $S$ is \emph{less than or equal to} a store $S'$ (written
$\leqstore{S}{S'}$) iff:
\begin{itemize}
\item $S' = \topS$, or
\item $\dom{S} \subseteq \dom{S'}$ and for all $l
\in \dom{S}$, $S(l) \userleq S'(l)$.
\end{itemize}

@@

\emph{By Definition 3:}, if $\userlub{d_1}{d_2} = \top$, then $\lubstore{\store{\storebindingRaw{l}{d_1}}}{\store{\storebindingRaw{l}{d_2}}} = \topS$. Notice that a store containing a binding $\storebindingRaw{l}{\top}$ can never arise during the execution of a $\lambdaLVar$ program, because an attempted write that would take the state of $l$ to $\top$ would raise an error before the write can occur.\\ \\

@@

An LVar’s state is now a pair $(d,frz )$, where $d$ is an element of the application-specific set $D$ and $frz$(Short for freeze) is a “status bit” of either $true$ or $false$. We define an ordering $\leqp$ on LVar states $\state{d}{\status}$ in terms of the application-specific ordering $\userleq$ on elements of $D$. Every element of $D$ is ``freezable'' except $\top$. Informally: 
\begin{itemize}
\item Two unfrozen states are ordered according to the application-specific $\userleq$; that is, $(d,false) \leqp (d',false)$
exactly when $d \userleq d'$.
\item Two frozen states do not have an order, unless they are equal: $(d,true) \leqp (d',true)$ exactly when $d = d'$.
\item An unfrozen state $(d,false)$ is less than or equal to a frozen state $(d',true)$ exactly when $d \userleq d'$.
\item A frozen state is less than an unfrozen state if the unfrozen state is $\top$; that is, $(d,true) \leqp (d',false)$ exactly when $d' = \top$.
\end{itemize}
The addition of status bits to the application-specific lattice results in a new lattice $(D_p, \leqp, \botp, \topp)$. $\lubp{}{}$ for the lub operation that $\leqp$ induces. During the evaluation of LVish programs, a \emph{store} $S$ keeps track of the states of LVars and represents them by a binding from a location $l$, drawn from a set $loc$, to its state, which is some pair $(d, frz)$ from the set $D_p$. For detailed syntax and semantics of Lvish please refer to the paper[7].\\ \\

\textbf{\textit{Communication primitives:}} They are same as Lvar except that new will now create a binding for a new Lvar whose initial state is (?, false). Put takes a pointer to the Lvar and a new lattice element which is a pair. Get similarly takes a threshold set containing pairs. $freeze\ e_{lv}\ after\ e_{events}\ with\ e_{cb}$ has the following semantics: It attaches the callback $e_{cb}$ to the LVar $e_{lv}$. The expression $e_{events}$ must evaluate to a event set $Q$; the callback will be executed, once, for each lattice element in $Q$ that the LVar’s state reaches or surpasses. The callback takes a lattice element as its argument. Its return value is ignored, so it runs solely for effect. For instance, a callback might itself do a put to the LVar to which it is attached, triggering yet more callbacks. If the handler reaches a quiescent state, the LVar $e_{lv}$ is frozen, and its exact state is returned (rather than an under-approximation of the state, as with $get$).

@@

\section{Implementation and Performance}
A prototype implementation of Lvish is available as a monadic library in Haskell at:\\ \\
\url{http://hackage.haskell.org/package/lvish}\\ \\
It adopts the basic approach of the par monad. The existing Par monad[3] in Haskell offers an API for parallel programming. It provides by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With
LVish we get a lattice-generic infrastructure: the Par monad itself, a thread scheduler, support for blocking and signaling threads, handler pools, and event handlers. This infrastructure does not guarantee quasi-determinism, only data structure
authors should import it who will implement a specific monotonic data structure using this library providing efficient parallel access, exporting a limited interface for its users ensuring quasi-determinism. LVish library provides its own custom scheduler and programs run only inside the par monad allowing only Lvish-sanctioned side effects to provide compile time guarantees of
determinism and quasi determinism.\\ \\
 In order to specify the type of the par computation it provides a phantom type that indicates the determinism level.\\ \\
\begin{lstlisting}
data Determinism = Det | QuasiDet
\end{lstlisting}
Par type constructor:
\begin{lstlisting}
Par : : Determinism -> * -> *
\end{lstlisting}
following suite of run functions\\ \\
\begin{lstlisting}
runPar : : Par Det a -> a\\
runParIO : : Par l v l a -> IO a
runParThenFreeze : : Deep frz a =>\\
	Par Det a -> FrzType a
\end{lstlisting}

If the Lvish code uses freeze then it is marked as QuasiDet or Det if it is fully deterministic. $runparIO$ allows us to execute Lvish code in the IO monad if the determinism level is arbitrary. $runparthenFreeze$s automatically freezes the Lvar when it returns back its exact value guaranteeing determinism.\\ \\
\textbf{\textit{Key Ideas:}} \textit{Atoms and Idempotence:} Atoms are elements not equal to $\bot$ but whose only smaller
element is $\bot$. Lattices for which every element is the lub of some set of atoms are called atomistic, and most application-specific lattices used by LVish programs have this property and the corresponding data structure usually exposes operations that
work at the atom level, semantically limiting puts to atoms, gets to threshold sets of atoms, and event sets to sets of atoms. A compact way to represent a change to the lattice is deltas allowing to easily and efficiently communicate such changes between
$puts$ and $gets$/handlers. For a set, a delta is an element; for a map, a key/value pair.\\
\textit{Idempotence,} meaning that $d\sqcup\d$ for any element $d$. Repeated puts or freezes have no eect,
if the scheduler is allowed to occasionally duplicate
work, it is possible to save on synchronization
costs. Since LVish computations are guaranteed
to be idempotent, we could use such a scheduler
(standard Chase-Lev deque [10]). Idempotence
helps to deal with races between put and get/ad-
dHandler. 
Representation: data LVars, data status, put, remove, listener, etc 
Internally par monad represents computations in continuation passing style, in terms of their interpretation in the IO monad. 
Signature and type of the following methods along with short explanation:
GetLV
putLV
freezeLV
Handler pools and quiescence

An example to illustrate the use of Lvish library is performing a parallel graph traversal given a starting node and finding all the nodes reachable from it. We first create a new LVar of a set type, called seen. Then we attach an event handler to the seen set. The newHandler function, defined elsewhere, takes 2 arguments, first is the LVar, seen, and second is the callback that we want to run every time an event arrives — that is, every time a new node is added to the set. The callback takes that newly arrived node node as its argument, looks up node’s neighbors in the graph by calling a function neighbors (also defined elsewhere), and then maps the insert function over the list of neighbor nodes that neighbors returns. Then, we add the starting node to the seen set with insert startNode seen and the event handler does the rest.
LVar operations can only run inside Par computations which are executed with one of the “run” functions.  runParThenFreeze, executes a Par computation and then returns a frozen LVar at the end of it, after an implicit global barrier. So, if we have a Par computation that side-effects an LVar and we run it with runParThenFreeze, then we don’t have to manually freeze the LVar ourselves, which means we don’t have to manually quiesce, either! 
This function returns a computation of type Par Det, for “deterministic”. The “Det” is an effect level, meaning that the computation can only perform deterministic operations — in particular, it’s not allowed to do a freeze, and if it tried to, then that would be a static type error. If we had called quiesce and freeze explicitly in our code within traverse function, then it would return a computation of type Par QuasiDet, for “quasi-deterministic”. 
traverse starts off the computation and then returns the seen LVar. The LVar may well still be growing at the time traverse returns — but, since runParThenFreeze cannot return unless all threads launched inside the Par computation are done, there’s no threat to determinism, and we’ll always get the same set of nodes as our answer. More generally, the fact that it’s possible to safely return an LVar while it is still being written to by a handler makes it possible to set up dataflow-style, pipeline-parallel networks of LVars, letting us combine task parallelism and pipeline parallelism in the same deterministic program.

\section{Summary}
\label{summary}

The summary shortly repeats the core ideas and results from the
previous text. If the reader has problems understanding the summary
he knows that he should go back to the relevant sections.
Thus, the last section should consist of:

\begin{itemize}
	\item a summary,
	\item an evaluation of what was done, importance of this work,
	\item what is left, what still needs to be done,
        \item short outlook into the future.
\end{itemize}

Last but not least, we can explain anything missing yet in the evaluation
done in this paper. This allows to refer to what readers can expect from
authors in the future.

% Put citations from bibtex into References section which were not
% explicity cited.
\nocite{robotron,
stonx,vice,650sim,herculessim,zib,4004,thermal1,thermal2,rojas,neumann,
neumann1void,neumann2void,neumann3,neumann4,Siewiorek,AmBlBr,Blaauw:1997:CAC,ChPB97,Brom98,Clym93,Buhu99,
Edwa01,Nill99,ABC+90,Rama91,Heid97,Kist95,Klot03,LMcCL92,LFS+92,LMDD92,Vlec03,Cray77,top500,650,Amda67,
Arla88,ODLD86,Hick04,Walk04}

>>>>

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically. \cite{lindsey}Written in a hypothetical parallel language. 
\begin{lstlisting}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
   _ = put lv 4  /write 4 in lv/
	  in v   /final value is v/
\end{lstlisting}
The \textit{let par} expression, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. \cite{lindsey}Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. \par 
\begin{lstlisting}
let par  _ = put l (4, )
         _ = put l (, 3) /error/
         in get l
\end{lstlisting}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

\textit{\textbf{Existing models}}\cite{kuper} A number of models currently exist as a part of the long running effort towards deterministic by construction parallel programming. \par

@@

In a directed graph where v represents a particular node, we wish to find all (or the first k degrees of) node connected to v, then analyze each node in that set. Existing parallel solutions\cite{kuper} often use a non-deterministic traversal of the connected component even though the final connected component is deterministic. For example, IVars cannot accumulate sets of visited nodes, nor can they be used as “mark bits” on visited nodes, since they can only be written once and not tested for emptiness. Streams, on the other hand, impose an excessively strict ordering for computing the unordered set of vertex labels in a connected component. A purely functional attempt is given in the original paper\cite{lkuper} where connected-component discovery is parallel, but members of the output set do not become available to other computations until component discovery is finished, limiting parallelism. 
An LVars\cite{lkuper} based programming model can possibly allow us to write a breadth first search traversal where threads communicate using shared monotonic data structures. Consumers of the data structure may execute as soon as data is available, but may only observe irrevocable, monotonic properties of it.

@@

To describe Lvar formally, a parallel call-by-value $\lambda$-calculus named $\lambdaLVar$ is introduced in the paper\cite{lkuper} which is extended with a $store$ and communication primitives $put$ and $get$ that operate on data in the $store$. The definition of $\lambdaLVar$ is parameterized by the choice of lattice and therefore $\lambdaLVar$ is actually a family of languages, rather than a single language. An LVar may have an arbitrary number of states forming a set $D$, which is partially ordered by a relation $\userleq$. Formally, $D$ is a \emph{bounded-join-semilattice} augmented with a greatest element $\top$.

% free floating figure using width of full page, to be put on [t]op.
\begin{figure}
\centerline{
\includegraphics[width=1.0\columnwidth]{Figures/lattice.jpg}
}
\caption{A nice caption. The larger width allows for more text without
taking too much space.}
\label{Fig2}
\end{figure}
\begin{itemize}
\item $D$ has a least element $\bot$, representing the initial “empty” state of an Lvar,
\item $D$ has a greatest element $\top$, representing the “error” state that results from conflicting updates to an Lvar,
\item $D$ comes equipped with a partial order $\userleq$, where $\bot \userleq d \userleq \top$ for all $d \in D$.
\item Every pair of elements in $D$ has a least upper bound (lub) written $\sqcup$ which means it is possible for two sub-computations to independently update an LVar, and then deterministically merge the results by taking the lub of the resulting two states.
\end{itemize}
Data structures like arrays, tree, maps , streams etc. can be represented  as lattice. $Figure$ $2$ gives three examples of lattices for common data structures.

During the evaluation of a $\lambdaLVar$ program, a \emph{store} $S$ keeps track of the states of LVars. Each LVar is represented by a \emph{binding} from a location $l$, drawn from a countable set $\Loc$, to its \emph{state}, which is some element $d \in D$. Although each LVar in a program has its own state, the states of all the LVars are drawn from the same lattice $D$. We can do this with no loss of generality because lattices corresponding to different types of LVars could always be union ed into a single lattice (with shared $\bot$ and $\top$ Elements). \\ \\
\textbf{Definition 1}: A \emph{store} is either a finite partial mapping $S : \Loc \fmap (D - \setof{\top})$, or the distinguished element $\topS$. \\ \\
The state space of $stores$ forms a \emph{bounded-join-semilattice} augmented with a greatest element, just as $D$ does, with the empty store $\bot_S$ as its least element and $\top_S$ as its greatest element. The $\userleq$ and $\sqcup$ operations defined on elements of $D$ can be lifted to the level of stores.\\ \\
\textbf{Definition 2}: 
 A store $S$ is \emph{less than or equal to} a store $S'$ (written
$\leqstore{S}{S'}$) iff:
\item $S' = \topS$, or
\item $\dom{S} \subseteq \dom{S'}$ and for all $l
\in \dom{S}$, $S(l) \userleq S'(l)$.

@@

\textbf{\textit{By Definition 3:}}, if $\userlub{d_1}{d_2} = \top$, then $\lubstore{\store{\storebindingRaw{l}{d_1}}}{\store{\storebindingRaw{l}{d_2}}} = \topS$. Notice that a store containing a binding $\storebindingRaw{l}{\top}$ can never arise during the execution of a $\lambdaLVar$ program, because an attempted write that would take the state of $l$ to $\top$ would raise an error before the write can occur.\\ \\

@@

An LVar’s state is now a pair $(d,frz )$, where $d$ is an element of the application-specific set $D$ and $frz$(Short for freeze) is a “status bit” of either $true$ or $false$. We define an ordering $\leqp$ on LVar states $\state{d}{\status}$ in terms of the application-specific ordering $\userleq$ on elements of $D$. Every element of $D$ is ``freezable'' except $\top$. Informally: 
\begin{itemize}
\item Two unfrozen states are ordered according to the application-specific $\userleq$; that is, $(d,false) \leqp (d',false)$
exactly when $d \userleq d'$.
\item Two frozen states do not have an order, unless they are equal: $(d,true) \leqp (d',true)$ exactly when $d = d'$.
\item An unfrozen state $(d,false)$ is less than or equal to a frozen state $(d',true)$ exactly when $d \userleq d'$.
\item A frozen state is less than an unfrozen state if the unfrozen state is $\top$; that is, $(d,true) \leqp (d',false)$ exactly when $d' = \top$.
\end{itemize}
The addition of status bits to the application-specific lattice results in a new lattice $(D_p, \leqp, \botp, \topp)$. $\lubp{}{}$ for the lub operation that $\leqp$ induces. During the evaluation of LVish programs, a \emph{store} $S$ keeps track of the states of LVars and represents them by a binding from a location $l$, drawn from a set $loc$, to its state, which is some pair $(d, frz)$ from the set $D_p$. For detailed syntax and semantics of Lvish please refer to the paper[7].\\ \\
\textbf{\textit{Communication primitives:}} They are same as Lvar except that new will now create a binding for a new Lvar whose initial state is (?, false). Put takes a pointer to the Lvar and a new lattice element which is a pair. Get similarly takes a threshold set containing pairs. $freeze\ e_{lv}\ after\ e_{events}\ with\ e_{cb}$ has the following semantics: It attaches the callback $e_{cb}$ to the LVar $e_{lv}$. The expression $e_{events}$ must evaluate to a event set $Q$; the callback will be executed, once, for each lattice element in $Q$ that the LVar’s state reaches or surpasses. The callback takes a lattice element as its argument. Its return value is ignored, so it runs solely for effect. For instance, a callback might itself do a put to the LVar to which it is attached, triggering yet more callbacks. If the handler reaches a quiescent state, the LVar $e_{lv}$ is frozen, and its exact state is returned (rather than an under-approximation of the state, as with $get$).

@@

\section{Implementation and Performance}
A prototype implementation of Lvish is available as a monadic library in Haskell at:\\ \\
\url{http://hackage.haskell.org/package/lvish}\\ \\
It adopts the basic approach of the par monad. The existing Par monad[3] in Haskell offers an API for parallel programming. It provides by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With
LVish we get a lattice-generic infrastructure: the Par monad itself, a thread scheduler, support for blocking and signaling threads, handler pools, and event handlers. This infrastructure does not guarantee quasi-determinism, only data structure
authors should import it who will implement a specific monotonic data structure using this library providing efficient parallel access, exporting a limited interface for its users ensuring quasi-determinism. LVish library provides its own custom scheduler and programs run only inside the par monad allowing only Lvish-sanctioned side effects to provide compile time guarantees of
determinism and quasi determinism.\\ \\
 In order to specify the type of the par computation it provides a phantom type that indicates the determinism level.\\ \\
data Determinism = Det | QuasiDet
Par type constructor:
Par : : Determinism -> * -> *
following suite of run functions\\ \\
runPar : : Par Det a -> a\\
runParIO : : Par l v l a -> IO a
runParThenFreeze : : Deep frz a =>\\
	Par Det a -> FrzType a
If the Lvish code uses freeze then it is marked as QuasiDet or Det if it is fully deterministic. $runparIO$ allows us to execute Lvish code in the IO monad if the determinism level is arbitrary. $runparthenFreeze$s automatically freezes the Lvar when it returns back its exact value guaranteeing determinism.\\ \\
\textbf{\textit{Key Ideas:}} \textit{Atoms and Idempotence:} Atoms are elements not equal to $\bot$ but whose only smaller
element is $\bot$. Lattices for which every element is the lub of some set of atoms are called atomistic, and most application-specific lattices used by LVish programs have this property and the corresponding data structure usually exposes operations that
work at the atom level, semantically limiting puts to atoms, gets to threshold sets of atoms, and event sets to sets of atoms. A compact way to represent a change to the lattice is deltas allowing to easily and efficiently communicate such changes between
$puts$ and $gets$/handlers. For a set, a delta is an element; for a map, a key/value pair.\par
\textit{Idempotence,} meaning that $d\sqcup{d=d}$ for any element $d$. Repeated puts or freezes have no effect, if the scheduler is allowed to occasionally duplicate work, it is possible to save on synchronization costs. Since LVish computations are guaranteed to be idempotent, we could use such a scheduler (standard Chase-Lev deque [10]). Idempotence helps to deal with races between $put$ and $get/addHandler$. \\ \\
Internally par monad represents computations in continuation passing style, in terms of their interpretation in the IO monad. The status bit of an LVar is tied together with a bag of waiting listeners, which include blocked gets and handlers.
data Status d = Frozen |
	Active (B.Bag(Listener d))
data LVar a d = LVar{state :: a,
	status :: IORef(Status d)} 
The bag module supports atomic insertion and removal, and concurrent traversal.\\
put :: Bag a -> a -> IO(Token a)
remove :: Token a -> IO()
foreach :: Bag a ->
   (a -> Token a -> IO()) -> IO()
A listener for an LVar is a pair of callbacks, one called when the LVars lattice value changes, and the other when the LVar is frozen. \\
data Listener d = Listener{
   onUpd :: d -> Token(Listener d)
   -> SchedQ -> IO(),
onFrz :: Token( Listener d)
   -> SchedQ -> IO() g
Internally, the Par monad represents computations in continuation-passing style, in terms of their interpretation in the IO monad.\\
type ClosedPar = SchedQ -> IO()
type ParCont a = a -> ClosedPar
mkPar :: (ParCont a -> ClosedPar)
   -> Par l v l a
The $ClosedPar$ type represents ready-to-run $Par$ computations, which are given direct access to the CPU-local scheduler queue. Rather than returning a final result, a completed $ClosedPar$ computation must call the scheduler, sched, on the queue. A $Par$
computation, on the other hand, completes by passing its intended result to its continuation yielding a $ClosedPar$ computation.
data HandlerPool = HandlerPool{
numCallbacks :: Counter,
blocked :: B.Bag ClosedPar}
The counter tracks the number of currently executing callbacks, which is used to implement quiesce. A bag of threads that are blocked waiting for the pool to reach a quiescent state. Please refer to[7] for more details on the library functions
including $quiesce$, $addhandler$ etc.\\ \\
\textbf{\textit{Communication primitives:}}
\item $getLV$ provides $get$ semantics. It takes a global and a delta threshold function as arguments.
\item $putLV$ provides $put$ semantics. Takes as argument an update function that performs $put$ on the underlying data structure. In situations where it races with a $freeze$, the $put$ happens first and then the LVar freeze status is checked.
To ensure correctness in such race situations,in addition to checking status bit, a mark bit is used in each CPU scheduler state.
\item $freezeLV$ provides freeze semantics. Takes an LVar as argument.
\textbf{\textit{Evaluation: k-CFA study}} In [7] a case study on k-CFA (parallelizing control flow analysis) is done to evaluate the performance of the Lvish implementation. K-CFA provide a hierarchy of increasingly precise methods to compute the control 
flow of a program or precisely the flow of values to expressions. The figure on the following page shows performance data[7] for LVish on 2 benchmarks. One is the blur benchmark based on k=2. The implementation using LVish for the 2 benchmarks uses lock-free data structures. Another synthetic benchmark is notChain which is simply a chain of 300 not functions that negates the benifit of the sharing approach in the LVish model.\\ \\
\textbf{\textit{Related work:}} In [6], feasibility of LVar threshold reads is explored in a distributed setting. Distributed systems involve replication of data objects across several locations making them robust to data loss and give better data locality. Eventual consistency is a policy in which the updates eventually reach the replicas and the replicas need not agree at all times. Convergent replicated data types provide a simple mathematical framework for reasoning about and enforcing the eventual consistency of such replicated objects, based on viewing them as elements of a lattice and replica conflict resolution as the lattice's join operation. CvRDTs are provably eventually consistent, but queries of CvRDTs allow inconsistent intermediate
states of replicas to be observed. LVar-style threshold queries to CvRDTs extend them with support for deterministic queries and show that they are strongly consistent queries. A threshold query that returns an answer when executed on a replica will return the same answer every subsequent time, and on a different replica, it will eventually return the same answer, otherwise will block until it does so. It is therefore impossible to observe different results from the same threshold query, whether at
different times on the same replica, or whether on different replicas.

% free floating figure using width of full page, to be put on [t]op.
\begin{figure}
\centerline{
\includegraphics[width=1.3\columnwidth]{Figures/speedup.jpg}
}
\caption{A nice caption. The larger width allows for more text without
taking too much space.}
\label{Fig2}
\end{figure}

\section{Summary}
\label{summary}
In an attempt towards creating better deterministic parallel programming models, Lvish has extended the existing usage of monotonicity as a theme to provide better applicability, usability, performance, providing a single generic model that subsumes
many of the single assignment models, thereby representing a larger class of problems which can be effciently parallelized. The performance with regards to certain benchmark applications are an evidence of the same. [4]discusses the viability of Lvish for certain graph algorithms, its drawbacks and proposed extension. Some of the drawbacks to highlight are the fact that Lvish side  effects are only monotonic updates and if a data structure needs to be updated multiple times it will have to be done out of
place thereby frequently allocating LVars becoming a source of additional memory allocation and ineffcient relative to in-place versions. Blocking $get$ semantics could slow down until the dependency is met reducing the performance, hence instead
of blocking, the tasks could be retried at a later time/rescheduled when there dependency has been satisfied. The overhead in going through the scheduler many times has also been highlighted in[4] for applications like Maximum independent set. The
paper proposes a solution similar to an abort/retry feature.
% Put citations from bibtex into References section which were not
% explicity cited.
\nocite{robotron,
stonx,vice,650sim,herculessim,zib,4004,thermal1,thermal2,rojas,neumann,
neumann1void,neumann2void,neumann3,neumann4,Siewiorek,AmBlBr,Blaauw:1997:CAC,ChPB97,Brom98,Clym93,Buhu99,
Edwa01,Nill99,ABC+90,Rama91,Heid97,Kist95,Klot03,LMcCL92,LFS+92,LMDD92,Vlec03,Cray77,top500,650,Amda67,
Arla88,ODLD86,Hick04,Walk04}
